Variation in use of East Asian Late Paleolithic weapons: A study of tip cross-sectional area of stemmed points from Korea

### 2022-11-09 analysis/paper/paper.qmd
Edited discussion

abstract: |
 The introduction of blade technology, stemmed points, end scrapers, burins, denticulates, and finer grained materials characterises the transition from the Early to Late Paleolithic in Korea. Stemmed points have been considered a representative tool that led this set of changes. In this research, we examine the possible role that stemmed points played during the technological transition, as well as throughout the Late Paleolithic. Our main questions are: What were stemmed points used for? How diverse were their likely uses? What are the temporal and spatial patterns in stemmed point use? We measured tip cross-sectional area (TCSA) to discriminate different likely use classes of projectile points, for example, as poisoned arrow tips or as stabbing spears. We analyzed TCSA with other variables including raw materials, weight, site and radiocarbon dates. Our results show that the stemmed points were likely used  as javelin tips and stabbing spear tips, with smaller numbers as dart tips and un-poisoned arrow tips. TCSA values are controlled mostly by size rather than raw material types. We found different usage of stemmed points at different sites, which could indicate people used stemmed points in different ways depending on the local environment. Some sites show a wide range of TCSA values that represent multi-purpose usage of stemmed points. The temporal pattern of TCSA values is one of little change throughout the Late Paleolithic period.<!-- SP were mainly used before Last Glacial Maximum (LGM). SP was mainly located in certain latitude. why? --> We conclude that stemmed points were  multi-functional tools, with many likely designed for javelin and stabbing spear tips.
keywords: |
  Stemmed points; Korean Paleolithic; Tip Cross-Sectional Area
---

@@

We summarized the distribution of TCSA values for each site to examine spatial patterns in TCSA values. We then explore possible environmental impact by connecting stemmed points on the map with eco-regional zones. We used QGIS 3.22.12 'Białowieża' (https://qgis.org/en/site/index.html) to draw a map based on the ecological study [CITE RESEARCH OF FINAL CHOICE] and used longitudes and latitudes of assemblages to locate the individual artifacts. Among `r n_sites` artifacts, we sat sites that contains fewer than 5 stemmed points aside and named them as "Other".

@@

```
We then grouped stemmed points into three categories based on LGM event: before, LGM, and after to examine possibility of environmental impact on the use of the artifacts (@fig-tcsa-radio-carbon-box). Our results shows that stemmed points were made predominantly before LGM and only a few left after LGM. Each category shows wide variation in LGM, which indicates diverse uses of stemmed points. There is no strong correlation between the TCSA values and the time period (p-value = `r p_lgm`).

@@

```{r}
#| label: fig-tcsa-sites
#| fig-cap: "Histograms of TCSA values by archaeological site."
#| fig-width:   4.5
#| fig-height:  4

@@

@@

veg_all <- read_sf(here("analysis/data/gis_data/eco_re_two.shp"))

@@

@@

for_plot <- ggplot() +
  geom_sf(data = veg_forest, aes(fill = "grey")) +
  geom_sf_text(data = veg_forest, aes(label = Zone), size =3.5)+
  geom_sf(data = pts_forest,size = 3, aes(color = min_tcsa)) +
  theme_minimal()+
  scale_fill_brewer(palette="Greys") +
  labs(color="TCSA") +
  guides(fill = FALSE) +
  theme(axis.text.x = element_text(size = 6)) +
  theme(axis.title.x = element_blank(),
          axis.title.y = element_blank())

@@

plot_grid(eco_plot, for_plot, eco_box, for_box, labels = c('A', 'C', 'B', 'D'))
```

Map A. @yim1975distribution, modified by @yi2010vegetation

Map B. @lee2008relationship

# Discussion ORDER OF TOPICS HERE MUST FOLLOW EXACTLY THE ORDER OF THE METHODS AND RESULTS SECTION

@@

Our results indicate that javelin tips and stabbing spear tips are the most probable ballistic uses for stemmed points. There are a few stemmed points in our sample that may have been used as dart tips and un-poisoned arrow tips, but none of them have TCSA values in the range of poison arrow tips (@fig-tcsa-all-sp). We found a wide range of TCSA values in our sample, indicating that stemmed points played a diverse role in forager toolkits. Some sites, such as Nosanri and Sinbuk, likely used stemmed points as restricted-function tools. Meanwhile, stemmed points were used for a wider set of tasks at other sites, especially Suyanggae. In general, stemmed points played multiple roles in more sites (@fig-tcsa-sites). Through the Late Paleolithic period, TCSA values showed little directional change (@fig-tcsa-radio-carbon). The chronological sequence of TCSA varies from region to region with no clear pattern. Most stemmed points appear to be have designed for use as javelin and stabbing spear tips, but there are also large numbers of tools that likely could fulfill multiple roles as armatures.

TCSA can, however, be impacted by other factors such as raw materials and portability. -@eren2022plains explain that smaller TCSA could be the result of pursuing production economy and transport efficiency. We explore the relationship between TCSA and raw material types and weight in order to examine these factors. Weight plays an important role in portability and penetration of projectiles [@hughes1998getting]. According to our results, raw materials do not have a clear pattern, but weight is correlated with raw materials (@fig-tcsa-raw-materials and @fig-tcsa-size). Different clusters of weight are matched with different types of weapon tips (@fig-artifact-size-cluster). Based on these analyses, we assume that people were able to produce the ideal size tools regardless of raw materials.

Our models premise that the choice of weapon-tip types reflects the socio-environmental circumstances. To examine the possible connections between the roles of stemmed points and population dynamics or environmental change during the Late Paleolithic period, we compared the temporal pattern of TCSA to results of Summed probability distribution (SPD) and simulated Mean Annual Temperature (MAT) from @park2022did. In those two analyses, -@park2022did covered assemblages with and without stemmed points, which is useful for us to explore the overall socio-environmental dynamics. As we discussed, there is no clear pattern in the temporal or regional pattern of TCSA (@fig-tcsa-radio-carbon or @no-figure-no-more: A). We consider that TCSA, or the function of the stemmed points, are strongly dependent on the surrounding environment or their situation. @no-figure-no-more: B shows that the temperature slowly decreased until around 22 ka and then went up again with a little range of fluctuations. Temperatures differ significantly between sites, with a maximum difference of 5℃. Results of SPD show several peaks, as indicated by the positive deviations from the three null models (shaded in red), and downs (shaded in blue), but no drastic growth or decline (@no-figure-no-more: C). From the climate and population study, we could observe that there is no dominating change but minor and often fluctuations throughout the Late Paleolithic and variations per site. Overall this pattern is well matched with the TCSA results, indicating the people produced stemmed points as multifunctional tools suitable for unexpected or various situations in their habitats similar to the case of Clovis points (@eren2022plains).

<--! In discussion we then compare those temporal patterns with demographic and environmental proxies based on summed probability distributions (SPD) and annual temperature changes from @park2022did. -->

>>>>

abstract: |
 The introduction of blade technology, stemmed points, end scrapers, burins, denticulates, and finer grained materials characterises the transition from the Early to Late Paleolithic in Korea. Stemmed points have been considered a representative tool that led this set of changes. In this research, we examine the possible role that stemmed points played during the technological transition, as well as throughout the Late Paleolithic. Our main questions are: What were stemmed points used for? How diverse were their likely uses? What are the temporal and spatial patterns in stemmed point use? We measured tip cross-sectional area (TCSA) to discriminate different likely use classes of projectile points, for example, as poisoned arrow tips or as stabbing spears. We analyzed TCSA with other variables including raw materials, weight, site and radiocarbon dates. Our results show that the stemmed points were likely used  as javelin tips and stabbing spear tips, with smaller numbers as dart tips and un-poisoned arrow tips. TCSA values are controlled mostly by size rather than raw material types. We found different usage of stemmed points at different sites, which could indicate people used stemmed points in different ways depending on the local environment. Some sites show a wide range of TCSA values that represent multi-purpose usage of stemmed points. The temporal pattern of TCSA values is one of little change throughout the Late Paleolithic period but SP were mainly used before Last Glacial Maximum (LGM). We observed that SP was located in certain ecoregions in Korea but no clear functional pattern is confirmed. We conclude that stemmed points were  multi-functional tools, with many likely designed for javelin and stabbing spear tips.

keywords: |
  Stemmed points; Korean Paleolithic; Tip Cross-Sectional Area
---

@@

We summarized the distribution of TCSA values for each assemblage to examine spatial patterns in TCSA values. We then explore possible environmental impact by connecting stemmed points on the map with eco-regional zones. We used QGIS 3.22.12 'Białowieża' (https://qgis.org/en/site/index.html) to draw a map based on the previous ecological studies [@yim1975distribution; @yi2010vegetation; @lee2008relationship] and located the individual artifacts using longitudes and latitudes of assemblages. Among `r n_sites` artifacts, we sat sites that contains fewer than 5 stemmed points aside and named them as "Other".

@@

We then grouped stemmed points into three categories based on LGM event: before, LGM, and after to examine possibility of environmental impact on the use of the artifacts (@fig-tcsa-radio-carbon-box). Our results shows that stemmed points were made predominantly before LGM and only a few left after LGM. Each category shows wide variation in LGM, which indicates diverse uses of stemmed points. There is no strong correlation between the TCSA values and the time period (p-value = `r p_lgm`).

@@

```{r}
#| label: fig-tcsa-per-sites
#| fig-cap: "Histograms of TCSA values by archaeological site."
#| fig-width:   4.5
#| fig-height:  4

@@

```{r}
#| label: spa-gis-tcsa-sites
#| fig-cap: "TCSA values by assemblages and ecoregions. A. Modern vegetation map defined by temporature and precipitation (modified the map from @yi2010vegetation, but original source is @yim1975distribution). This map include 3 zones. Central Temperature Zone (CT), South Temperature Zone (ST), and Subtropical-warm Temperature Zone (SW). The individual points indicate assemblages. B. Boxplots for  vegitation zones. The individual points indicate artifacts. C. Ecoregion map based on geographical boundaries and ecological conditions (modified the map from @lee2008relationship). The individual points indicate assemblages. This map include 14 zones. Imjin river basin (IRB), Metropolitan (MP), Central inland (CI), Kangwon coastal (KC), Choongnam coastal (CC), Southwestern inland (SWI), Upper Nagdong river basin (UNRB), WoolYoung coastal (WYC), Western Cholla (WC), Southern mountain (SM), Southeastern inland (SEI), Hyungsan Taewha coastal (HTC), Western south coastal (WSC), and Eastern south coastal (ESC). The individual points indicate assemblages. D. Boxplots for  ecoregion zones, excludings zones with no artifacts. The individual points indicate artifacts."
#| fig-width:   5
#| fig-height:  4

@@

library(sf)
veg_all <- read_sf(here("analysis/data/gis_data/eco_re_two.shp"))

@@

eco_lgm <- round(summary(eco_aov)[[1]][[5]][1], digits = 3)

@@

for_plot <- ggplot() +
  geom_sf(data = veg_forest, aes(fill = "grey")) +
  geom_sf_text(data = veg_forest, aes(label = Zone), size =3.5)+
  geom_sf(data = pts_forest, aes(color = min_tcsa)) +
  theme_minimal()+
  scale_fill_brewer(palette="Greys") +
  labs(color="TCSA") +
  guides(fill = FALSE) +
  theme(axis.text.x = element_text(size = 6)) +
  theme(axis.title.x = element_blank(),
          axis.title.y = element_blank())

@@

for_lgm <- round(summary(for_aov)[[1]][[5]][1], digits = 3)

plot_grid(eco_plot, for_plot, eco_box, for_box, labels = c('A', 'C', 'B', 'D'))
```
We then located individual assemblages on maps that show ecoregional boundaries to explore spatial patterns. We used two maps from the previous studies that represents different zones based on vegetation and geography [@yi2010vegetation; @yim1975distribution; @lee2008relationship]. Our results show that stemmed points are only located in certain zones, Central Temperature Zone (CT) or Central inland (CI) (@spa-gis-tcsa-sites), for example. We observed that TCSA values are stonely accociated with the zones (p-value for Vegetation zones = `r eco_lgm`, p-value for Ecoregion zones = `r for_lgm`). However, there is no clear functional pattern throughout the zones (@spa-gis-tcsa-sites panel B and D). There is a wide variation in TCSA regardless of the stemmed points density in each zone.

# Discussion

@@

Our results indicate that javelin tips and stabbing spear tips are the most probable ballistic uses for stemmed points. There are a few stemmed points in our sample that may have been used as dart tips and un-poisoned arrow tips, but none of them have TCSA values in the range of poison arrow tips (@fig-tcsa-all-sp). But in general, we found a wide range of TCSA in our sample, indicating that stemmed points played a diverse role in forager toolkits. TCSA can be impacted by other factors such as raw materials and portability. -@eren2022plains explain that smaller TCSA could be the result of pursuing production economy and transport efficiency. We explore the relationship between TCSA and raw material types and weight in order to examine these factors. We observed no clear pattern for raw materials (@fig-tcsa-raw-materials) but weight (@fig-artifact-size-cluster). Interestingly, weight is correlated with raw materials (@fig-tcsa-size). Therefore we could assume that raw materials also associated with TCSA value. We found that different clusters of weight are matched with different types of weapon tips (@fig-artifact-size-cluster). The TCSA varies from assemblages to assemblages without any discernible temporal pattern in functional use (@fig-tcsa-radio-carbon). But the stemmed points mainly used before LGM (@fig-tcsa-radio-carbon-box). We found that stemmed points are only located in certain ecoregional zones such as Central Temperature zone without any clear pattern in function (@spa-gis-tcsa-sites). Both temporal and spatial analyses show that stemmed points were used for a particular environment, not so much harsh as LGM.

Our models premise that the choice of weapon-tip types reflects the socio-environmental circumstances. To examine the possible connections between the roles of stemmed points and population dynamics or environmental change during the Late Paleolithic period, we compared the temporal and spatial patterns of TCSA to results of simulated Mean Annual Temperature (MAT) and Summed probability distribution (SPD) from @park2022did. We found clear correlation between stemmed points and temperature (@spa-gis-tcsa-sites and @fig-tcsa-radio-carbon-box, and Figure 9 from @park2022did)
around 4 to 6 ℃

 We consider that TCSA, or the function of the stemmed points, are strongly dependent on the surrounding environment or their situation. @no-figure-no-more: B shows that the temperature slowly decreased until around 22 ka and then went up again with a little range of fluctuations. Temperatures differ significantly between sites, with a maximum difference of 5℃. Results of SPD show several peaks, as indicated by the positive deviations from the three null models (shaded in red), and downs (shaded in blue), but no drastic growth or decline (@no-figure-no-more: C). From the climate and population study, we could observe that there is no dominating change but minor and often fluctuations throughout the Late Paleolithic and variations per site. Overall this pattern is well matched with the TCSA results, indicating the people produced stemmed points as multifunctional tools suitable for unexpected or various situations in their habitats similar to the case of Clovis points (@eren2022plains).

\<--! In discussion we then compare those temporal patterns with demographic and environmental proxies based on summed probability distributions (SPD) and annual temperature changes from @park2022did. --\>
